---
title: "select_best_scROSMAPsubclusters"
format: html
editor: visual
---

```{r}
library(Seurat)
library(edgeR)
library(NMOF)
library(parallel)
library(tidyverse)
library(deltaccd)
setwd("~/Box Sync/Henry_stuff/AD_project/PseudoBulkVillas/Jan12/")
load("rosmap424_pseudobulk_by_cluster.rda")
```

Read in ROSMAP metadata

```{r}
#clean up column (subcluster names)
colnames(counts) = str_replace(colnames(counts), "_donor", "")
colnames(counts) = str_replace(colnames(counts), ".rds", "")
rownames(pseudobulk_meta) = colnames(counts)

#read in subject metadata
rosmap_subject_metadata = read.csv("~/Box Sync/Henry_stuff/AD_project/scROSMAP/Meta_data/cleaned_rosmap_meta_cogdxConds.csv")

#read in another file with sequencing info for all scRNAseq done for ROSMAP
rosmap_scassay_meta = read_csv("~/Box Sync/Henry_stuff/AD_project/human_data/ROSMAP/ExpandedROSMAP/MetaData/ROSMAP_assay_scrnaSeq_metadata.csv")

#Seurat now complains when groub.by argument in AggregrateCounts is numeric so I'll change them in rosmap_metadata and pseudobulk_meta:
pseudobulk_meta$project_id = paste0("X", pseudobulk_meta$project_id)
rosmap_subject_metadata$projid = paste0("X", rosmap_subject_metadata$projid)

#remove the cogdx 6 subjects from the metadata file
rosmap_meta = filter(rosmap_subject_metadata, cogdx != 6)

#merge large meta_data file with sequencing info
rosmap_scassay_meta = rosmap_scassay_meta %>% separate_wider_delim(cols = 2, delim = "_", names = c("col1", "individualID"), too_many = "drop", too_few = "align_start")
# full_meta = merge(rosmap_meta, rosmap_scassay_meta, by = "individualID", all.x = T)
rosmap_meta$batch = rosmap_scassay_meta$rnaBatch[match(rosmap_meta$individualID, rosmap_scassay_meta$individualID)]
rosmap_meta$batch = str_extract(rosmap_meta$batch, "(?<=-B)[0-9]*") %>% as.numeric
rosmap_meta = rosmap_meta %>%
  mutate(batch_bin = cut(batch, c(4, 10, 64), right = F))

##### AUGMENT ROSMAP_META #####
rosmap_meta = rosmap_meta %>%
  mutate(braaksc_bin = cut(braaksc, c(0, 3, 5, 7), right = F))
rosmap_meta = rosmap_meta %>%
  mutate(ceradsc_bin = cut(ceradsc, c(1, 3, 5), right = F))
rosmap_meta$apoe_ordinal  = 1
rosmap_meta$apoe_ordinal[rosmap_meta$apoe_genotype == 34 | rosmap_meta$apoe_genotype == 24] = 2
rosmap_meta$apoe_ordinal[rosmap_meta$apoe_genotype == 44 ] = 3
rosmap_meta$apoe_ordinal[is.na(rosmap_meta$apoe_genotype) ] = NA
rosmap_meta$age_death_numeric = rosmap_meta$age_death %>% str_replace("\\+", "") %>% as.numeric

#adds cogdx score for each subcluster
pseudobulk_meta$cogdx = rosmap_subject_metadata$cogdx[match(pseudobulk_meta$project_id, rosmap_subject_metadata$projid)]
```

Create Seurat object

```{r}
#creates a seurat object
seurat_obj = CreateSeuratObject(
  counts,
  project = "scRNA ROSMAP",
  assay = "RNA",
  names.field = 3,
  names.delim = "_",
  meta.data = pseudobulk_meta)

refCor = getRefCor(species = "human", tissue = "pan", useEntrezGeneId = F)
```

Loss function for grid search:

```{r}

loss = function(str){
  expr = FetchData(seurat_obj, vars = "cluster")
  tmp = seurat_obj[,grep(str, expr$cluster)]
  
  cts <- AggregateExpression(tmp,
                             group.by = c("project_id"),
                             assays = 'RNA',
                             slot = "counts",
                             return.seurat = FALSE)
  cts = cts$RNA[,na.exclude(match(rosmap_meta$projid, colnames(cts$RNA)))]
  my_rosmap_meta = rosmap_meta[match(colnames(cts), rosmap_meta$projid), ]
  

  #create DGElist
  y <- DGEList(counts = cts, group = my_rosmap_meta$cond_cogdx)
  #default params for filterByExpression
  keep.genes = filterByExpr(y)
  if (all(keep.genes[rownames(refCor)])){
  #filter out the genes
  y = y[keep.genes, , keep=FALSE]
  # print(paste("keeping", dim(y)[1], "genes"))
  
  my_tmm_DGE_obj <- calcNormFactors(y, method = "TMM")
  my_tmm = cpm(my_tmm_DGE_obj) #calculate tmms: https://www.biostars.org/p/317701/
  my_tmm = as.data.frame(my_tmm)
  

  # all(colnames(my_tmm) == (rosmap_meta$projid))
  try({
  ans = calcCCD(refCor, emat = as.data.frame(my_tmm[rownames(refCor),]), groupVec = my_rosmap_meta$cond_cogdx)
  return(ans$CCD[1])},
  silent = T
  
  )
  return(Inf)
  } 
  return(Inf)
  }

```

```{r}
n = 3
l <- rep(list(0:1), n)
df=expand.grid(l)
colnames(df) = paste0("Mic\\.", 1:3,"$")
#colnames(df) = paste0("Ast\\.", 1:10,"$")

strs = c()
for (i in 2:nrow(df)){
  out = paste(colnames(df)[which(df[i,] == 1)], collapse = "|")
  #out = str_replace_all(out, "\\.", "\\\\.")
  strs = c(strs, out)
}

```

```{r}
tic = Sys.time()
grid_results = gridSearch(loss, levels = list(str = strs), method = 'multicore', mc.control = list(mc.cores = 10))
toc = Sys.time()
print(toc-tic)
# Exc.3|Exc.5
# or Exc.3|Exc.5|Exc.10
#or "Ast\\.1$|Ast\\.2$|Ast\\.5$|Ast\\.6$|Ast\\.7$|Ast\\.8$"
```

take the best subtypes and make a tmm

```{r}
get_subluster_data = function(str){
  expr = FetchData(seurat_obj, vars = "cluster")
  tmp = seurat_obj[,grep(str, expr$cluster)]
  
  cts <- AggregateExpression(tmp,
                             group.by = c("project_id"),
                             assays = 'RNA',
                             slot = "counts",
                             return.seurat = FALSE)
  cts = cts$RNA[,na.exclude(match(rosmap_meta$projid, colnames(cts$RNA)))]
}
cts = get_subluster_data("Inh\\.2$|Inh\\.8$|Inh\\.11$|Inh\\.12$|Inh\\.13$")
#"Inh\\.2$|Inh\\.8$|Inh\\.11$|Inh\\.12$|Inh\\.13$"
#"Exc\\.5$|Exc\\.3$"
# "Ast\\.1$|Ast\\.2$|Ast\\.5$|Ast\\.6$|Ast\\.7$|Ast\\.8$"
# "Mic\\.4$|Mic\\.5$|Mic\\.6$|Mic\\.7$|Mic\\.8$|Mic\\.10$|Mic\\.11$|Mic\\.13$|Mic\\.15$|Mic\\.16$"
my_rosmap_meta = rosmap_meta[match(colnames(cts), rosmap_meta$projid), ]


#create DGElist
y <- DGEList(counts = cts, group = my_rosmap_meta$cond_cogdx)
#default params for filterByExpression
keep.genes = filterByExpr(y)

keep.genes[rownames(refCor)] = T ##TODO Watch out for this line

#filter out the genes
y = y[keep.genes, , keep=FALSE]

my_tmm_DGE_obj <- calcNormFactors(y, method = "TMM")
my_tmm = cpm(my_tmm_DGE_obj) #calculate tmms: https://www.biostars.org/p/317701/
my_tmm = as.data.frame(my_tmm)
plotHeatmap(rownames(refCor), emat = as.data.frame(my_tmm), groupVec = my_rosmap_meta$cond_cogdx)

# rm_vars1 = rowSums(my_tmm[,my_rosmap_meta$cond_cogdx])
# rm_vars2 = rowSums(my_tmm[,!my_rosmap_meta$cond_cogdx])
# rm_rows = which(rm_vars1 == 0 | rm_vars2 == 0)
# keep_refCor = which(rownames(refCor) %in% rownames(my_tmm[-rm_rows,]))

# calcCCD(refCor[keep_refCor, keep_refCor], emat = as.data.frame(my_tmm[-rm_rows,]), groupVec = my_rosmap_meta$cond_cogdx)

# calcDeltaCCD(refCor[keep_refCor, keep_refCor], as.data.frame(my_tmm[-rm_rows,]), groupVec = my_rosmap_meta$cond_cogdx, groupNormal = 0)

calcCCD(refCor, emat = as.data.frame(my_tmm), groupVec = my_rosmap_meta$cond_cogdx)

calcDeltaCCD(refCor, as.data.frame(my_tmm), groupVec = my_rosmap_meta$cond_cogdx, groupNormal = 0)






```

The correlagrams of the Exc.3 and Exc.5 by themselves is bad. When you add them, the correlagram is great. Therefore the clusters could be in part based on the expression of a rhythmic gene. Below I will see that:

```{r}
cts3 = get_subluster_data("Exc\\.3$")
cts5 = get_subluster_data("Exc\\.5$")

sapply(rownames(refCor), function(x){
  max1 = max(c(cts3[x,], cts5[x,]))
  min1 = min(c(cts3[x,], cts5[x,]))
  breaks = seq(max1, min1, length.out = 20)
  c1 <- rgb(173,216,230,max = 255, alpha = 90, names = "lt.blue")
  c2 <- rgb(255,192,203, max = 255, alpha = 90, names = "lt.pink")
  h1 = hist(cts3[x,], plot = F, breaks = breaks, freq = F)
  h2 = hist(cts5[x,], plot = F, breaks = breaks, freq = F)
  max_y = max(c(h1$density, h2$density))
  plot(h1, col = c1, main = x)
  plot(h2, col = c2, add = T)
  # boxplot(cts3[x,], cts5[x,], main = x)
})
```

Write out tmm

```{r}
setwd("~/Box Sync/Henry_stuff/AD_project/human_data/Cyclops_folders/tmms/scROSMAP/cogdx_controls/EdgeR_filt_normed/")
out = rownames_to_column(my_tmm, var = "Gene_Symbols")
covs = c("Cond_D", paste0("cond_", my_rosmap_meta$cond))
covs1 = c("Batch_D", paste0("cond_", my_rosmap_meta$batch_bin))
covs1 = str_replace(covs1, "\\[10,64\\)", "0")
covs1 = str_replace(covs1, "\\[4,10\\)", "1")
covs2 = c("ceradsc_D", paste0("cond_",  my_rosmap_meta$ceradsc_bin))
covs3 = c("braaksc_D", paste0("cond_",  my_rosmap_meta$braaksc_bin))

out = rbind(covs, covs1, covs2, covs3, out)
colnames(out) = str_replace(colnames(out), "X", "")
write.table(out, "InhNeurons_2_8_11_12_13_FiltByExprDefault_TMM.csv", sep = ',', row.names = F)

#write.table(my_rosmap_meta, "~/Box Sync/Henry_stuff/AD_project/scROSMAP/Meta_data/rosmap_meta_Exc3and5.csv", sep = ",", row.names = F)
```

PCA the tmm

```{r}
pc_res = prcomp(t(my_tmm), scale. = T)
colorvec = my_rosmap_meta$batch_bin
ggplot(as.data.frame(pc_res$x), aes(x = PC1, y = PC2, color = colorvec))+geom_point()
ggplot(as.data.frame(pc_res$x), aes(x = PC2, y = PC3, color = colorvec))+geom_point()
ggplot(as.data.frame(pc_res$x), aes(x = PC1, y = PC3, color = colorvec))+geom_point()
```

COMBAT the TMM

```{r}
library(sva)
#double check metadata is in order
all(my_rosmap_meta$projid== colnames(cts)) 

adjusted_counts <- ComBat_seq(as.matrix(cts), batch=my_rosmap_meta$batch_bin, group=my_rosmap_meta$cond_cogdx, covar_mod=NULL, full_mod=TRUE,shrink=FALSE, shrink.disp=FALSE, gene.subset.n=NULL)
```

```{r}
all(colnames(cts) == (my_rosmap_meta$projid))

#create DGElist
y_combat <- DGEList(counts = adjusted_counts, group = my_rosmap_meta$cond_cogdx)
#default params for filterByExpression
keep.genes = filterByExpr(y_combat)
#filter out the genes
y_combat = y_combat[keep.genes, , keep=FALSE]
print(paste("keeping", dim(y_combat)[1], "genes"))

my_tmm_DGE_obj <- calcNormFactors(y_combat, method = "TMM")
my_tmm_combat = cpm(my_tmm_DGE_obj) #calculate tmms: https://www.biostars.org/p/317701/
my_tmm_combat = as.data.frame(my_tmm_combat)

all(colnames(my_tmm_combat) == (my_rosmap_meta$projid))
```

Write out combated tmm

```{r}
setwd("~/Box Sync/Henry_stuff/AD_project/human_data/Cyclops_folders/tmms/scROSMAP/cogdx_controls/EdgeR_filt_normed/")
out = rownames_to_column(my_tmm_combat, var = "Gene_Symbols")
covs = c("Cond_D", paste0("cond_", my_rosmap_meta$cond))
covs1 = c("Batch_D", paste0("cond_", my_rosmap_meta$batch_bin))
covs1 = str_replace(covs1, "\\[10,64\\)", "0")
covs1 = str_replace(covs1, "\\[4,10\\)", "1")
covs2 = c("ceradsc_D", paste0("cond_",  my_rosmap_meta$ceradsc_bin))
covs3 = c("braaksc_D", paste0("cond_",  my_rosmap_meta$braaksc_bin))

out = rbind(covs, covs1, covs2, covs3, out)
colnames(out) = str_replace(colnames(out), "X", "")
write.table(out, "InhNeurons_2_8_11_12_13_FiltByExprDefault_TMM_CombatAdjusted.csv", sep = ',', row.names = F)
```

```{r}
pc_res = prcomp(t(my_tmm_combat), scale. = T)
colorvec = my_rosmap_meta$batch_bin
ggplot(as.data.frame(pc_res$x), aes(x = PC1, y = PC2, color = colorvec))+geom_point()
ggplot(as.data.frame(pc_res$x), aes(x = PC2, y = PC3, color = colorvec))+geom_point()
ggplot(as.data.frame(pc_res$x), aes(x = PC1, y = PC3, color = colorvec))+geom_point()
```

### edgeR Differential Expression

```{r}
y_combat = estimateDisp(y_combat)
```

```{r}
et <- exactTest(y_combat, pair = c(1,2))
topTags(et)
```

```{r}
design <- model.matrix(~y$samples$group)
fit <- glmQLFit(y_combat, design)
qlf.2vs1 <- glmQLFTest(fit, coef=2)
topTags(qlf.2vs1)
```

### Write out DE results

```{r}
setwd("~/Box Sync/Henry_stuff/AD_project/scROSMAP/simple_differential_expr/")

#write out results from the exact test
et$table$FDR = p.adjust(et$table$PValue, method = "BH")
write.table(et$table, "Mglia_4_5_6_7_8_10_11_13_15_16/ Microglia_4_5_6_7_8_10_11_13_15_16_filtByExpr_combatSeq_edgeR_DE_exactTest.csv", sep =',', row.names = T, col.names = NA )

#write out qlf test
qlf.2vs1$table$FDR = p.adjust(qlf.2vs1$table$PValue, method = "BH")

write.table(qlf.2vs1$table, "Mglia_4_5_6_7_8_10_11_13_15_16/ Microglia_4_5_6_7_8_10_11_13_15_16_filtByExpr_combatSeq_edgeR_DE_QLFTest.csv", sep =',', row.names = T, col.names = NA )

```
