---
title: "select_best_scROSMAPsubclusters"
format: html
editor: visual
---

```{r}
library(Seurat)
library(edgeR)
library(NMOF)
library(parallel)
library(tidyverse)
library(deltaccd)
library(gridExtra)
load("../rosmap424_pseudobulk_by_cluster.rda")
```

### Read in ROSMAP metadata

```{r}
#clean up column (subcluster names)
colnames(counts) = str_replace(colnames(counts), "_donor", "")
colnames(counts) = str_replace(colnames(counts), ".rds", "")
rownames(pseudobulk_meta) = colnames(counts)

#read in subject metadata
rosmap_subject_metadata = read.csv("../ROSMAP_metadata/cleaned_rosmap_meta_cogdxConds.csv")

#read in another file with sequencing info for all scRNAseq done for ROSMAP
rosmap_scassay_meta = read_csv("../ROSMAP_metadata/ROSMAP_assay_scrnaSeq_metadata.csv", show_col_types = F)

#Seurat now complains when groub.by argument in AggregrateCounts is numeric so I'll change them in rosmap_metadata and pseudobulk_meta:
pseudobulk_meta$project_id = paste0("X", pseudobulk_meta$project_id)
rosmap_subject_metadata$projid = paste0("X", rosmap_subject_metadata$projid)

#remove the cogdx 6 subjects from the metadata file
rosmap_meta = filter(rosmap_subject_metadata, cogdx != 6)

#merge large meta_data file with sequencing info
rosmap_scassay_meta = rosmap_scassay_meta %>% separate_wider_delim(cols = 2, delim = "_", names = c("col1", "individualID"), too_many = "drop", too_few = "align_start")
# full_meta = merge(rosmap_meta, rosmap_scassay_meta, by = "individualID", all.x = T)
rosmap_meta$batch = rosmap_scassay_meta$rnaBatch[match(rosmap_meta$individualID, rosmap_scassay_meta$individualID)]
rosmap_meta$batch = str_extract(rosmap_meta$batch, "(?<=-B)[0-9]*") %>% as.numeric
rosmap_meta = rosmap_meta %>%
  mutate(batch_bin = cut(batch, c(4, 10, 64), right = F))

##### AUGMENT ROSMAP_META #####
rosmap_meta = rosmap_meta %>%
  mutate(braaksc_bin = cut(braaksc, c(0, 3, 5, 7), right = F))
rosmap_meta = rosmap_meta %>%
  mutate(ceradsc_bin = cut(ceradsc, c(1, 3, 5), right = F))
rosmap_meta$apoe_ordinal  = 1
rosmap_meta$apoe_ordinal[rosmap_meta$apoe_genotype == 34 | rosmap_meta$apoe_genotype == 24] = 2
rosmap_meta$apoe_ordinal[rosmap_meta$apoe_genotype == 44 ] = 3
rosmap_meta$apoe_ordinal[is.na(rosmap_meta$apoe_genotype) ] = NA
rosmap_meta$age_death_numeric = rosmap_meta$age_death %>% str_replace("\\+", "") %>% as.numeric

#adds cogdx score for each subcluster
pseudobulk_meta$cogdx = rosmap_meta$cogdx[match(pseudobulk_meta$project_id, rosmap_meta$projid)]

pseudobulk_meta$cond_cogdx = rosmap_meta$cond_cogdx[match(pseudobulk_meta$project_id, rosmap_meta$projid)]

pseudobulk_meta$pmi = rosmap_meta$pmi[match(pseudobulk_meta$project_id, rosmap_meta$projid)]

pseudobulk_meta$sex = rosmap_meta$msex[match(pseudobulk_meta$project_id, rosmap_meta$projid)]

sex_table = table(rosmap_meta$cond_cogdx, rosmap_meta$msex)
colnames(sex_table) = c("female", "male")
rownames(sex_table)= c("cond_0", "cond_1")
```

### Create Seurat object

notes: "NA" on these plots means the projectid from the 424 .rda file is not found in my rosmap_meta. These are the cogdx 3,6 subjects I exclude.

```{r}
#creates a seurat object
seurat_obj = CreateSeuratObject(
  counts,
  project = "scRNA ROSMAP",
  assay = "RNA",
  names.field = 3,
  names.delim = "_",
  meta.data = pseudobulk_meta)

#get Ribo Protein percent
seurat_obj <- PercentageFeatureSet(seurat_obj, "^RP[SL]", col.name = "percent_ribo")

# Percentage hemoglobin genes - includes all genes starting with HB except HBP.
seurat_obj <- PercentageFeatureSet(seurat_obj, "^HB[^(P)]", col.name = "percent_hb")

#Does number of ribosoal features (%ribo * n_features) change with cogdx
RPSL.genes <- grep(pattern = "^RP[SL]", x = rownames(x = seurat_obj@assays$RNA$counts), value = TRUE)

counts.RPS <- Matrix::colSums(seurat_obj@assays$RNA$counts[RPSL.genes, ])

seurat_obj = AddMetaData(
  object = seurat_obj,
  metadata = counts.RPS,
  col.name = "cnts_ribo"
)
VlnPlot(seurat_obj, group.by = "cond_cogdx", features = c("cnts_ribo", "pmi"), pt.size = 0.1, ncol = 1, alpha = 0)

#plot QC mertrics
feats <- c("nFeature_RNA", "nCount_RNA", "percent_ribo")
VlnPlot(seurat_obj, group.by = "broadclass", features = feats, pt.size = 0.1, ncol = 2, alpha = 0.05) +
  NoLegend()

#creates a reference correlation matrix
refCor = getRefCor(species = "human", tissue = "pan", useEntrezGeneId = F)
```

### Loss function for grid search:

```{r}

loss = function(str){
  expr = FetchData(seurat_obj, vars = "cluster")
  tmp = seurat_obj[,grep(str, expr$cluster)]
  
  cts <- AggregateExpression(tmp,
                             group.by = c("project_id"),
                             assays = 'RNA',
                             slot = "counts",
                             return.seurat = FALSE)
  cts = cts$RNA[,na.exclude(match(rosmap_meta$projid, colnames(cts$RNA)))]
  my_rosmap_meta = rosmap_meta[match(colnames(cts), rosmap_meta$projid), ]
  

  #create DGElist
  y <- DGEList(counts = cts, group = my_rosmap_meta$cond_cogdx)
  #default params for filterByExpression
  keep.genes = filterByExpr(y)
  if (all(keep.genes[rownames(refCor)])){
  #filter out the genes
  y = y[keep.genes, , keep=FALSE]
  # print(paste("keeping", dim(y)[1], "genes"))
  
  my_tmm_DGE_obj <- calcNormFactors(y, method = "TMM")
  my_tmm = cpm(my_tmm_DGE_obj) #calculate tmms: https://www.biostars.org/p/317701/
  my_tmm = as.data.frame(my_tmm)
  

  # all(colnames(my_tmm) == (rosmap_meta$projid))
  try({
  ans = calcCCD(refCor, emat = as.data.frame(my_tmm[rownames(refCor),]), groupVec = my_rosmap_meta$cond_cogdx)
  return(ans$CCD[1])},
  silent = T
  
  )
  return(Inf)
  } 
  return(Inf)
  }

```

### Generate combinations for testing ccd

```{r}
n = 10 #number of subclusters to consider
l <- rep(list(0:1), n)
df=expand.grid(l)
colnames(df) = paste0("Ast\\.", 1:n,"$") #replace `Ast` for other subclusters, ie. Exc, Inh, Mic, etc.

strs = c()
for (i in 2:nrow(df)){
  out = paste(colnames(df)[which(df[i,] == 1)], collapse = "|")
  strs = c(strs, out)
}

```

### Run Grid Search

```{r}
tic = Sys.time()
grid_results = gridSearch(loss, levels = list(str = strs), method = 'multicore', mc.control = list(mc.cores = 10))
toc = Sys.time()
print(toc-tic)
# Exc.3|Exc.5
#"Ast\\.1$|Ast\\.2$|Ast\\.5$|Ast\\.6$|Ast\\.7$|Ast\\.8$"
```

### Take the best subtypes and make a tmm

```{r}
get_subluster_data = function(str){
  expr = FetchData(seurat_obj, vars = "cluster")
  tmp = seurat_obj[,grep(str, expr$cluster)]
  
  cts <- AggregateExpression(tmp,
                             group.by = c("project_id"),
                             assays = 'RNA',
                             slot = "counts",
                             return.seurat = FALSE)
  cts = cts$RNA[,na.exclude(match(rosmap_meta$projid, colnames(cts$RNA)))]
}
cts = get_subluster_data(".*")
#"Inh\\.2$|Inh\\.8$|Inh\\.11$|Inh\\.12$|Inh\\.13$"
#"Exc\\.5$|Exc\\.3$"
# "Ast\\.1$|Ast\\.2$|Ast\\.5$|Ast\\.6$|Ast\\.7$|Ast\\.8$"
# "Mic\\.4$|Mic\\.5$|Mic\\.6$|Mic\\.7$|Mic\\.8$|Mic\\.10$|Mic\\.11$|Mic\\.13$|Mic\\.15$|Mic\\.16$"
my_rosmap_meta = rosmap_meta[match(colnames(cts), rosmap_meta$projid), ]


#create DGElist
y <- DGEList(counts = cts, group = my_rosmap_meta$cond_cogdx)
#default params for filterByExpression
keep.genes = filterByExpr(y)

keep.genes[rownames(refCor)] = T #make sure genes needed for cormat are kept

#filter out the genes
y = y[keep.genes, , keep=FALSE]

my_tmm_DGE_obj <- calcNormFactors(y, method = "TMM")
my_tmm = cpm(my_tmm_DGE_obj) #calculate tmms: https://www.biostars.org/p/317701/
my_tmm = as.data.frame(my_tmm)
plotHeatmap(rownames(refCor), emat = as.data.frame(my_tmm), groupVec = my_rosmap_meta$cond_cogdx)

# rm_vars1 = rowSums(my_tmm[,my_rosmap_meta$cond_cogdx])
# rm_vars2 = rowSums(my_tmm[,!my_rosmap_meta$cond_cogdx])
# rm_rows = which(rm_vars1 == 0 | rm_vars2 == 0)
# keep_refCor = which(rownames(refCor) %in% rownames(my_tmm[-rm_rows,]))

# calcCCD(refCor[keep_refCor, keep_refCor], emat = as.data.frame(my_tmm[-rm_rows,]), groupVec = my_rosmap_meta$cond_cogdx)

# calcDeltaCCD(refCor[keep_refCor, keep_refCor], as.data.frame(my_tmm[-rm_rows,]), groupVec = my_rosmap_meta$cond_cogdx, groupNormal = 0)

calcCCD(refCor, emat = as.data.frame(my_tmm), groupVec = my_rosmap_meta$cond_cogdx)

calcDeltaCCD(refCor, as.data.frame(my_tmm), groupVec = my_rosmap_meta$cond_cogdx, groupNormal = 0)


```

The correlagrams of the Exc.3 and Exc.5 by themselves is bad. When you add them, the correlagram is great. Therefore the clusters could be in part based on the expression of a rhythmic gene. Below I will see that:

```{r}
cts3 = get_subluster_data("Exc\\.3$")
cts5 = get_subluster_data("Exc\\.5$")

sapply(rownames(refCor), function(x){
  max1 = max(c(cts3[x,], cts5[x,]))
  min1 = min(c(cts3[x,], cts5[x,]))
  breaks = seq(max1, min1, length.out = 20)
  c1 <- rgb(173,216,230,max = 255, alpha = 90, names = "lt.blue")
  c2 <- rgb(255,192,203, max = 255, alpha = 90, names = "lt.pink")
  h1 = hist(cts3[x,], plot = F, breaks = breaks)
  h2 = hist(cts5[x,], plot = F, breaks = breaks)
  max_y = max(c(h1$density, h2$density))
  plot(h1, col = c1, main = x)
  plot(h2, col = c2, add = T)
  # boxplot(cts3[x,], cts5[x,], main = x)
})
```

### Write out tmm

```{r}
setwd("../normed_counts/")
out = rownames_to_column(my_tmm, var = "Gene_Symbols")
covs = c("Cond_D", paste0("cond_", my_rosmap_meta$cond))
covs1 = c("Batch_D", paste0("cond_", my_rosmap_meta$batch_bin))
covs1 = str_replace(covs1, "\\[10,64\\)", "0")
covs1 = str_replace(covs1, "\\[4,10\\)", "1")
covs2 = c("ceradsc_D", paste0("cond_",  my_rosmap_meta$ceradsc_bin))
covs3 = c("braaksc_D", paste0("cond_",  my_rosmap_meta$braaksc_bin))
covs4 = c("sex_D", paste0("cond_", my_rosmap_meta$msex))
#fill one missing pmi value to median of pmi's:
my_rosmap_meta$pmi[which(is.na(my_rosmap_meta$pmi))] = median(my_rosmap_meta$pmi, na.rm = T)
covs5 = c("pmi_C", my_rosmap_meta$pmi)

out = rbind(covs, covs1, covs2, covs3, covs4, covs5, out)
colnames(out) = str_replace(colnames(out), "X", "")
#Change filename depending on what cell type is selected
write.table(out, "InhNeurons_2_8_11_12_13_FiltByExprDefault_TMM.csv", sep = ',', row.names = F)

```

### PCA the tmm

```{r}
CYCLOPS_seed_genes = c("DBP", "TEF", "ZNF580", "GFOD1", "CASTOR2", "ETV5", "FMNL1", "MAMLD1", "TM7SF2", "FPGS", "SYNJ2", "HLF", "KBTBD2", "MPP2", "MRGBP", "USP2", "CC2D1B", "GALNT9", "GSTT2B", "NR1D1", "SRSF5", "ZMAT3", "ARNTL", "FAM3A", "FOSL2", "HUNK", "MORN1", "NEXN", "SPHK2", "TRIM9", "ENC1", "GUF1", "IGFBP4", "PER2", "CRY1", "MRAS", "SFPQ", "ARRDC1", "CFAP46", "DNAJB1", "MKKS", "SARS2", "SEC14L1", "SETD4", "SLC6A6", "SPATA7", "TFR2", "PER1", "POR", "SLC16A1", "PER3", "NPAS2", "TSC22D3", "BHLHE41", "NR1D2", "LEO1", "STK35", "GRAMD4", "COQ10B", "DTX4", "PDZRN3", "DUSP11", "OPRL1", "LRRC39", "TRIM24", "RARA", "LDB1", "BHLHE40", "RSPO2", "TRAF5", "ANKRD12")
CYCLOPS_seed_genes = CYCLOPS_seed_genes[which(CYCLOPS_seed_genes %in% rownames(my_tmm))]
pc_res = prcomp(t(my_tmm), scale. = T)
pc_res_seeds = prcomp(t(my_tmm[CYCLOPS_seed_genes,]), scale. = T)


plot_pcs = function(pc_res, colorvec, title, continuous = F){
 if(!continuous){colorvec = as.factor(colorvec)}
  p1 = ggplot(as.data.frame(pc_res$x), aes(x = PC1, y = PC2, color = colorvec))+geom_point() + {if(continuous)scale_color_gradient(low = "blue", high = "red")}
  p2 = ggplot(as.data.frame(pc_res$x), aes(x = PC2, y = PC3, color = colorvec))+geom_point()+ {if(continuous)scale_color_gradient(low = "blue", high = "red")}
  p3 = ggplot(as.data.frame(pc_res$x), aes(x = PC1, y = PC3, color = colorvec))+geom_point()+ {if(continuous)scale_color_gradient(low = "blue", high = "red")}
  p4 = ggplot(as.data.frame(pc_res$x), aes(x = PC1, y = PC4, color = colorvec))+geom_point()+ {if(continuous)scale_color_gradient(low = "blue", high = "red")}
  p5 = ggplot(as.data.frame(pc_res$x), aes(x = PC2, y = PC4, color = colorvec))+geom_point()+ {if(continuous)scale_color_gradient(low = "blue", high = "red")}
  p6 = ggplot(as.data.frame(pc_res$x), aes(x = PC3, y = PC4, color = colorvec))+geom_point()+ {if(continuous)scale_color_gradient(low = "blue", high = "red")}
  grid.arrange(p1, p2, p3,p4,p5, p6, nrow = 3, top=textGrob(title,gp=gpar(fontsize=12,font=3)))
}

plot_pcs(pc_res, my_rosmap_meta$batch_bin, "PCA All Genes, Seq. Batch")
plot_pcs(pc_res_seeds, my_rosmap_meta$batch_bin, "PCA CYCLOPS Seed Genes, Seq. Batch")
plot_pcs(pc_res, my_rosmap_meta$msex, "PCA All Genes, Sex")
plot_pcs(pc_res_seeds, my_rosmap_meta$msex, "PCA CYCLOPS Seed Genes, Sex")
plot_pcs(pc_res, my_rosmap_meta$pmi, "PCA All Genes, PMI", continuous = T)
plot_pcs(pc_res_seeds, my_rosmap_meta$pmi, "PCA CYCLOPS Seed Genes, PMI", continuous = T)
```

### COMBAT the TMM

```{r}
library(sva)
#double check metadata is in order
all(my_rosmap_meta$projid== colnames(cts)) 

adjusted_counts <- ComBat_seq(as.matrix(cts), batch=my_rosmap_meta$batch_bin, group=my_rosmap_meta$cond_cogdx, covar_mod=NULL, full_mod=TRUE,shrink=FALSE, shrink.disp=FALSE, gene.subset.n=NULL)
```

```{r}
all(colnames(cts) == (my_rosmap_meta$projid))

#create DGElist
y_combat <- DGEList(counts = adjusted_counts, group = my_rosmap_meta$cond_cogdx)
#default params for filterByExpression
keep.genes = filterByExpr(y_combat)
#filter out the genes
y_combat = y_combat[keep.genes, , keep=FALSE]
print(paste("keeping", dim(y_combat)[1], "genes"))

my_tmm_DGE_obj <- calcNormFactors(y_combat, method = "TMM")
my_tmm_combat = cpm(my_tmm_DGE_obj) #calculate tmms: https://www.biostars.org/p/317701/
my_tmm_combat = as.data.frame(my_tmm_combat)

all(colnames(my_tmm_combat) == (my_rosmap_meta$projid))
```

### Write out combated tmm

```{r}
setwd("../normed_counts/")
out = rownames_to_column(my_tmm_combat, var = "Gene_Symbols")
covs = c("Cond_D", paste0("cond_", my_rosmap_meta$cond))
covs1 = c("Batch_D", paste0("cond_", my_rosmap_meta$batch_bin))
covs1 = str_replace(covs1, "\\[10,64\\)", "0")
covs1 = str_replace(covs1, "\\[4,10\\)", "1")
covs2 = c("ceradsc_D", paste0("cond_",  my_rosmap_meta$ceradsc_bin))
covs3 = c("braaksc_D", paste0("cond_",  my_rosmap_meta$braaksc_bin))
covs4 = c("sex_D", paste0("cond_", my_rosmap_meta$msex))
#fill one missing pmi value to median of pmi's:
my_rosmap_meta$pmi[which(is.na(my_rosmap_meta$pmi))] = median(my_rosmap_meta$pmi, na.rm = T)
covs5 = c("pmi_C", my_rosmap_meta$pmi)

out = rbind(covs, covs1, covs2, covs3, covs4, covs5, out)
colnames(out) = str_replace(colnames(out), "X", "")
write.table(out, "InhNeurons_2_8_11_12_13_FiltByExprDefault_TMM_CombatAdjusted.csv", sep = ',', row.names = F)
```

### PCA on the combated TMM

```{r}

CYCLOPS_seed_genes = CYCLOPS_seed_genes[which(CYCLOPS_seed_genes %in% rownames(my_tmm_combat))]

pc_res = prcomp(t(my_tmm_combat), scale. = T)
pc_res_seeds = prcomp(t(my_tmm_combat[CYCLOPS_seed_genes,]), scale. = T)


plot_pcs = function(pc_res, colorvec, title, continuous = F){
 if(!continuous){colorvec = as.factor(colorvec)}
  p1 = ggplot(as.data.frame(pc_res$x), aes(x = PC1, y = PC2, color = colorvec))+geom_point() + {if(continuous)scale_color_gradient(low = "blue", high = "red")}
  p2 = ggplot(as.data.frame(pc_res$x), aes(x = PC2, y = PC3, color = colorvec))+geom_point()+ {if(continuous)scale_color_gradient(low = "blue", high = "red")}
  p3 = ggplot(as.data.frame(pc_res$x), aes(x = PC1, y = PC3, color = colorvec))+geom_point()+ {if(continuous)scale_color_gradient(low = "blue", high = "red")}
  p4 = ggplot(as.data.frame(pc_res$x), aes(x = PC1, y = PC4, color = colorvec))+geom_point()+ {if(continuous)scale_color_gradient(low = "blue", high = "red")}
  p5 = ggplot(as.data.frame(pc_res$x), aes(x = PC2, y = PC4, color = colorvec))+geom_point()+ {if(continuous)scale_color_gradient(low = "blue", high = "red")}
  p6 = ggplot(as.data.frame(pc_res$x), aes(x = PC3, y = PC4, color = colorvec))+geom_point()+ {if(continuous)scale_color_gradient(low = "blue", high = "red")}
  grid.arrange(p1, p2, p3,p4,p5, p6, nrow = 3, top=textGrob(title,gp=gpar(fontsize=12,font=3)))
}

plot_pcs(pc_res, my_rosmap_meta$batch_bin, "PCA All Genes, Seq. Batch")
plot_pcs(pc_res_seeds, my_rosmap_meta$batch_bin, "PCA CYCLOPS Seed Genes, Seq. Batch")
plot_pcs(pc_res, my_rosmap_meta$msex, "PCA All Genes, Sex")
plot_pcs(pc_res_seeds, my_rosmap_meta$msex, "PCA CYCLOPS Seed Genes, Sex")
plot_pcs(pc_res, my_rosmap_meta$pmi, "PCA All Genes, PMI", continuous = T)
plot_pcs(pc_res_seeds, my_rosmap_meta$pmi, "PCA CYCLOPS Seed Genes, PMI", continuous = T)
```

### edgeR Differential Expression

```{r}
y_combat = estimateDisp(y_combat)
```

```{r}
et <- exactTest(y_combat, pair = c(1,2))
topTags(et)
```

```{r}
design <- model.matrix(~y$samples$group)
fit <- glmQLFit(y_combat, design)
qlf.2vs1 <- glmQLFTest(fit, coef=2)
topTags(qlf.2vs1)
```

### Write out DE results

```{r}
setwd("../edgeR_diff_expression/")

#write out results from the exact test
et$table$FDR = p.adjust(et$table$PValue, method = "BH")
write.table(et$table, "Inhib_2_8_11_12_13/ Inhib_2_8_11_12_13_filtByExpr_combatSeq_edgeR_DE_exactTest.csv", sep =',', row.names = T, col.names = NA )

#write out qlf test
qlf.2vs1$table$FDR = p.adjust(qlf.2vs1$table$PValue, method = "BH")

write.table(qlf.2vs1$table, "Inhib_2_8_11_12_13/ Inhib_2_8_11_12_13_filtByExpr_combatSeq_edgeR_DE_QLFTest.csv", sep =',', row.names = T, col.names = NA )

```

Alternative - no combat, supply edgeR with batch information:

```{r}
all(colnames(cts) == (my_rosmap_meta$projid))

#create DGElist
y <- DGEList(counts = as.matrix(cts), group = my_rosmap_meta$cond_cogdx)
#default params for filterByExpression
keep.genes = filterByExpr(y)
#filter out the genes
y = y[keep.genes, , keep=FALSE]
print(paste("keeping", dim(y)[1], "genes"))

design = model.matrix(~my_rosmap_meta$cond_cogdx + my_rosmap_meta$batch_bin)
colnames(design) = c("CTL","AD", "Batch10_64")

y1 = estimateDisp(y, design)
fit <- glmQLFit(y1, design)
qlf <- glmQLFTest(fit, coef=2)
topTags(qlf)
```
